Synthesis of Linear Functional Programs

### 2021-05-07 article/article.tex
merge linear logic background

\mypara{Goals}
 % Goals
In the end, we intend to be able to do full and partial synthesis of
well-typed programs. Full synthesis consists of the production of a
function (or set of) satisfying the specification; Partial synthesis
is the ``completion'' of a partial program (i.e. a function with a
\emph{hole} in it). The work will start from a small core linear --
i.e. \emph{resource-aware} functional language, building up to
recursive types/functions, with potential other avenues of further
extension. 
%
We will evaluate the work through
expressiveness benchmarks (i.e. can we synthesize ``X''?) and time
measurements (i.e. how fast can we synthesize ``X''?').

% \begin{enumerate}
% \item General motivation
%   \begin{itemize}
% \item 
% \item Specifications appear in many forms\dots
% \item Program synthesis as a means of improving program
%   correctness/programmer productivity.
% \item Challenges: space of valid programs, space of valid
%   specifications,  etc.
% \end{itemize}
% \item \emph{Type-based} program synthesis
%   \begin{itemize}
%  \item Rich types as specifications.
% \item Languages with richer type systems allow for more precise
%   types. This precision can statically eliminate various kinds of
%   logical errors by making certain invalid program states ill-typed
%   (e.g. a ``null aware'' type system, will ensure you cannot
%   dereference a null-pointer).
% \item But can also be a burden -- the whole point of these type
%   systems is to ensure that ``less'' things are well-typed, and
%   sometimes it's hard to convince the checker\dots
% \item Type-based synthesis leverages rich types as a way of pruning
%   the search space and by using types as a more ``familiar''
%   specification.
% \end{itemize}
% \item ``My'' problem: Synthesis based on linear types
%   \begin{itemize}
%   \item Linear types constrain resource usage in programs by \emph{statically} limiting
%     the number of times certain resources can be used during its
%     lifetime.
%   \item Applications in resource-aware programming such as concurrent
%     programming (e.g. session types) and memory-management
%     (e.g. Rust).
%   \item This work will study program synthesis in a functional,
%     linearly-typed setting, by leveraging the propositions-as-types
%     correspondence between linear logic and the (linear)
%     $\lambda$-calculus.
%   \item What is props-as-types? What is linear logic? Linear $\lambda$
%     is ``just'' a small functional core language.
%   \item Through props-as-types, synthesis can be formulated as proof
%     search and so we can leverage a lot of related technology from the
%     literature.
%   \end{itemize}
% \item Goals:
%   \begin{itemize}
% \item Full and partial synthesis of well-typed (in the above sense)
%   programs. Full synthesis means producing a function (or set of)
%   satisfying the spec. Partial synthesis means being given a partial
%   program (i.e. a function with a hole in it) and ``completing it''.
% \item Start from a small system, build up to recursive
%   types/functions.
% \item Potential to go further: type refinements; interactive
%   synthesis?
%  \item Evaluation through expressiveness benchmarks (can we synthesize
%    ``X''?); time measurements (how fast can we synthesize ``X''?');
%    etc.
%   \end{itemize}
% \end{enumerate}

@@

% Linear Logic
\mypara{Linear Logic}

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.

>>>>

\mypara{Goals}
 % Goals
In the end, we intend to be able to do full and partial synthesis of
well-typed programs. Full synthesis consists of the production of a
function (or set of) satisfying the specification; partial synthesis
is the ``completion'' of a partial program (i.e. a function with a
\emph{hole} in it). The work will start from a small core linear --
i.e. \emph{resource-aware} functional language, building up to
recursive types/functions, with potential other avenues of further
extension. 
%
We will evaluate the work through
expressiveness benchmarks (i.e. can we synthesize ``X''?) and time
measurements (i.e. how fast can we synthesize ``X''?').

@@

\mypara{Linear Logic}

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes into linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.
To present the formalization of this logic, besides the new connectives, we need to introduce the \emph{resource-aware context} $\Delta$.
 In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types, but in which each and every variable must be used exactly once.
 So, to introduce the connective $\tensor$ which defines a multiplicative pair of propositions, we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
 to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
it useable more than once. To distinguish the ``infinite'' variables, a separate context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
